---
title: "Advent of Code: Day 3"
author: "Irene Steves"
date: "February 16, 2018"
output: github_document
---

Compared to [Day 3](https://adventofcode.com/2017/day/3), Days 1 and 2 were a breeze! It definitely put that growing coding ego in check.

## Part I

```
You come across an experimental new kind of memory stored on an infinite two-dimensional
grid.

Each square on the grid is allocated in a spiral pattern starting at a location marked 
1 and then counting up while spiraling outward. For example, the first few squares are
allocated like this:

17  16  15  14  13
18   5   4   3  12
19   6   1   2  11
20   7   8   9  10
21  22  23---> ...
While this is very space-efficient (no squares are skipped), requested data must be
carried back to square 1 (the location of the only access port for this memory system)
by programs that can only move up, down, left, or right. They always take the shortest
path: the Manhattan Distance between the location of the data and square 1.

For example:

Data from square 1 is carried 0 steps, since it's at the access port.
Data from square 12 is carried 3 steps, such as: down, left, left.
Data from square 23 is carried only 2 steps: up twice.
Data from square 1024 must be carried 31 steps.
How many steps are required to carry the data from the square identified in your puzzle
input all the way to the access port?
```

After chasing down a few dead ends, I found patterns in the squares that helped me figure out a (round-about) way to determine the final position:

#### 1. The bottom-right corner of the square is always an the square of an odd number.

```
x   x   x   x   x   x   x
x   x   x   x   x   x   x
x   x   x   1   x   x   x
x   x   x   x   9   x   x
x   x   x   x   x   25  x
x   x   x   x   x   x   49

Square 1: 1^2 = 1
Square 2: 3^2 = 9
Square 3: 5^2 = 25
Square 4: 7^2 = 49
```

In math language, this means that for each square `n`, the bottom-right value is $(2n - 1)^2$.

#### 2. The number of digits in each square is the difference between the values at the bottom right corners of (a) the square of interest and (b) the next square inside.

```
Square 1: 1 - 0 = 1
Square 2: 9 - 1 = 8
Square 3: 25 - 9 = 16
Square 4: 49 - 25 = 24
```

Math translation: $(2n - 1)^2 - (2n - 3)^2$.

With the exception of Square 1, the number of digits in each square is always divisible by 4. In fact, our mathematical expression can be simplified into: $8(n - 1)$.

#### 3. With each successive square, the corners are two steps further away from the center.

Since we care primarily about position, we can re-write our spiral from above as steps from the center (zero):

```
6   5   4   3   4   5   6
5   4   3   2   3   4   5
4   3   2   1   2   3   4
3   2   1   0   1   2   3
4   3   2   1   2   3   4
5   4   3   2   3   4   5
6   5   4   3   4  -->
```

Looking at the bottom-right (or any) corner, you can see that it gains two steps with each square:

```
Square 1: 0 steps
Square 2: 2 steps
Square 3: 4 steps
Square 4: 6 steps
```

In math: $2(n - 1)$




If we write it out for each square, then we get:

```
Square 1: 0
Square 2: 1 2 1 2 1 2 1 2
Square 3: 3 2 3 4 3 2 3 4 3 2 3 4 3 2 3 4
Square 4: 5 4 3 4 5 6 5 4 3 4 5 6 5 4 3 4 5 6 5 4 3 4 5 6
```

From this, we see that:

- Each square (after 1) ends with an even number, specfically, $2(n - 1)$
- Each square (after 1) starts with an odd number, specifically, one less than the even number: $2(n - 1) - 1$ = $2n - 3$
- The minimum number of steps in each square is equal to $n - 1$
- Except for square 1, the number of digits in each square is divisible by 4 (since there are four sides in a square)

```
Square 1: 0
Square 2: (1 2) (1 2) (1 2) (1 2)
Square 3: (3 2 3 4) (3 2 3 4) (3 2 3 4) (3 2 3 4)
Square 4: (5 4 3 4 5 6) (5 4 3 4 5 6) (5 4 3 4 5 6) (5 4 3 4 5 6)
```

Now, we can start pulling some of these patterns together:

```{r warning = FALSE, message = FALSE}
library(tidyverse)
steps <- tibble(square = 1:10) %>% 
    mutate(max_steps = 2*(square - 1),
           min_steps = square - 1,
           start_steps = max_steps - 1,
           end_steps = max_steps) 
print(steps)
```

Except for the first squares, we know that for each side, the number of steps starts at some number (`start_steps`), decreases by one until it gets to the minimum number of steps (`min_steps`), and then increases by one until it gets to the end, or maximum, number of steps (`end_steps` or `max_steps`). Repeat this four times and we have the whole sequence of steps for each square. 


```{r warning = FALSE, message = FALSE}
positions <- steps %>% 
    rowwise() %>% 
    mutate(seq = paste(c(start_steps:min_steps, (min_steps + 1):end_steps), 
                       collapse = " "),
           seq4 = paste(rep(seq, 4), collapse = " ")) %>% 
    mutate(seq4 = replace(seq4, square == 1, 0)) %>%  #replace square 1's sequence with 0
    select(seq4)
print(positions)
```

If you now paste it all together as a long string (with a tiny bit of base R), and then separate out again into a vector of integers, we can answer the puzzle question!

```{r warning = FALSE, message = FALSE}
str_c(positions$seq4, collapse = " ") %>% 
    str_split(" ", simplify = TRUE) %>% 
    as.numeric()

```

We can't quite get to the puzzle answer with just the first 10 squares, so we'll have to expand our scope a little. Let's pull everything together in one giant pipe:

```{r warning = FALSE, message = FALSE}
positions500 <- tibble(square = 1:500) %>% 
    mutate(max_steps = 2*(square - 1),
           min_steps = square - 1,
           start_steps = max_steps - 1,
           end_steps = max_steps) %>% 
    
    rowwise() %>% 
    mutate(seq = paste(c(start_steps:min_steps, (min_steps + 1):end_steps), 
                       collapse = " "),
           seq4 = paste(rep(seq, 4), collapse = " ")) %>% 
    mutate(seq4 = replace(seq4, square == 1, 0)) %>%  #replace square 1's sequence with 0
    select(seq4) %>% 
    
    str_c(collapse = " ") %>% 
    #directly using str_c on the tibble, rather than tibble$seq4 
    #results in the inclusion of some special characters
    #so, rather than str_split at all the spaces,
    #we can extract all digits instead:
    str_extract_all("\\d+", simplify = TRUE) %>% 
    as.numeric()
```

Let's run some tests to make sure it works!

```{r warning = FALSE, message = FALSE}
positions500[1] == 0
positions500[12] == 3
positions500[23] == 2
positions500[1024] == 31
```
