# Stream Processing

## Part I

On [Day 9](https://adventofcode.com/2017/day/9), you sit on a stream bank and see a stream of characters filled with "garbage" that you need to parse out. Here's some of the syntax rules paraphrased:

- **groups** begin with { and end with }
- within a group, there are 0 or more things separated by commas
    - the "things" can be other groups or garbage
- } closes the most-recently-opened unclosed group (nesting relationships are possible)
- **garbage** begins with < and ends with >
    - anything can appear within a piece of garbage, including { or } or <
- ! cancels out the next character, which includes <, > or even !

Some examples of garbage:

```
<>, empty garbage.
<random characters>, garbage containing random characters.
<<<<>, because the extra < are ignored.
<{!>}>, because the first > is canceled.
<!!>, because the second ! is canceled, allowing the > to terminate the garbage.
<!!!>>, because the second ! and the first > are canceled.
<{o"i!a,<{i<a>, which ends at the first >.
```

Examples of groups:

```
{}, 1 group.
{{{}}}, 3 groups.
{{},{}}, also 3 groups.
{{{},{},{{}}}}, 6 groups.
{<{},{},{{}}>}, 1 group (which itself contains garbage).
{<a>,<a>,<a>,<a>}, 1 group.
{{<a>},{<a>},{<a>},{<a>}}, 5 groups.
{{<!>},{<!>},{<!>},{<a>}}, 2 groups (since all but the last > are canceled).
```

Each group is also worth a different number of points depending on how deeply nested it is. The outermost group gets 1 point, and the next group in gets 2 points, and so on. Example scores include:

```
{}, score of 1.
{{{}}}, score of 1 + 2 + 3 = 6.
{{},{}}, score of 1 + 2 + 2 = 5.
{{{},{},{{}}}}, score of 1 + 2 + 3 + 3 + 3 + 4 = 16.
{<a>,<a>,<a>,<a>}, score of 1.
{{<ab>},{<ab>},{<ab>},{<ab>}}, score of 1 + 2 + 2 + 2 + 2 = 9.
{{<!!>},{<!!>},{<!!>},{<!!>}}, score of 1 + 2 + 2 + 2 + 2 = 9.
{{<a!>},{<a!>},{<a!>},{<ab>}}, score of 1 + 2 = 3.
```

The goal of this exercise is to determine the total score for all groups in the input. Let's start with a test string.

```{r}
#test string
string <- "{{<!>},{<!>},{<!>},{<a>}}"
```

Since `!` cancels out the next character, no matter what it is, we'll start with that. We can use [regular expressions](https://www.rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf).

```{r}
string <- str_replace_all(string, "!\\D", "")
string
```

To remove the garbage, we can go through our string iteratively and grab the first `<` and first `>` that we find. Once there are no `>` left, we're done with our cleanup process.

```{r}
while(str_detect(string, ">") == TRUE){
    garbage_start <- str_locate(string, "<")[1] #first < in seq
    garbage_end <- str_locate(string, ">")[1] #first > in seq
    str_sub(string, garbage_start, garbage_end) <- "" #remove garbage
}
```

For the last step, we need to add up the points to get our final "score" for the stream of strings. We can set up a `summary` data frame to calculate the number of points each `{` receives. We define positive points (`pospoints`) as any `{` encountered before the `{` of interest. Any `}` encountered before the `{` of interest then contributes negative points. 

```{r}
#counting {}
#count { based on position; for every } encountered, minus 1
summary <- data.frame(locs = str_locate_all(string, "\\{")[[1]][,1],
                      pospoints = NA,
                      negpoints = NA)

summary <- summary %>% 
    as.tibble() %>% 
    rowwise() %>% 
    mutate(pospoints = str_count(substr(string, 1, locs), "\\{"),
           negpoints = str_count(substr(string, 1, locs), "\\}"),
           points = pospoints - negpoints)

sum(summary$points)
```

With an easy `sum` function, we can calculate the total score for the stream of strings!

## Part II

In this section, you have to "prove" you removed the garbage by counting what you removed.

```
The leading and trailing < and > don't count, nor do any canceled characters or the ! doing the canceling.

<>, 0 characters.
<random characters>, 17 characters.
<<<<>, 3 characters.
<{!>}>, 2 characters.
<!!>, 0 characters.
<!!!>>, 0 characters.
<{o"i!a,<{i<a>, 10 characters.

How many non-canceled characters are within the garbage in your puzzle input?
```

Let's start again with the test string, and remove all cancelled letters. We can then define a new vector `n_garbage` to record the number of characters in each piece of garbage. Once we finish with the process, we just have to sum it up!

```{r}
string <- "{{<!>},{<!>},{<!>},{<a>}}" %>% 
    str_replace_all("!\\D", "")

n_garbage <- NULL
while(str_detect(string, ">") == TRUE){
    garbage_start <- str_locate(string, "<")[1]
    garbage_end <- str_locate(string, ">")[1]
    n_garbage <- c(n_garbage,
                   nchar(str_sub(string, garbage_start, garbage_end)) - 2)
    #-2 because < and > don't count
    
    str_sub(string, garbage_start, garbage_end) <- ""
}

sum(n_garbage)
```

